Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-02-05 ms.tex
final fix to app D

We distinguish between small, intermediate, or large tumors.
In small ($N\ll N_a^*\ll N_m^*$) and intermediate ($N_a^* \ll N \ll N_m^*$) tumors, we have at most one lineage that rescues the cancer cell population.
As a result, the recurrence time is given by~\citep{avanzini2019cancer}
\begin{align}\label{meanproliferationtime}
\tau_a^r&\approx\tau_a+\frac{\log \left(p_mN\right)}{\Delta_m}.
\end{align}
The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}. % change 35 
In both scenarios, plugging \cref{meantimet2} for $\tau_a$ in \cref{meanproliferationtime} provides a good approximation for the recurrence time (black line in \Cref{ProliferationTimeLarge}).
For small tumors, we have a simpler approximation by plugging \cref{limitapprox4} for $\tau_a$ to get $\tau_a^r\approx1/|\Delta_a| + \log{\left(p_mN\right)}/\Delta_m$ (blue line in \Cref{ProliferationTimeLarge}).

In a large tumor, $N_m^* \ll N$, the sensitive population produces a large number of rescue lineages in a short period of time.
As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}

>>>>

We distinguish between small, intermediate, or large tumors.
In small ($N\ll N_a^*\ll N_m^*$) tumors, we have at most one lineage that rescues the cancer cell population.
As a result, the recurrence time is given by~\citep{avanzini2019cancer}
\begin{align}\label{meanproliferationtime}
\tau_a^r&\approx\tau_a+\frac{\log \left(p_mN\right)}{\Delta_m}.
\end{align}
The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}. % change 35 
Therefore, in small tumors, we can substitue \cref{limitapprox4} for $\tau_a$ in \cref{meanproliferationtime} to get (blue line in \Cref{ProliferationTimeLarge})
\begin{equation*}
\tau_a^r\approx1/|\Delta_a| + \log{\left(p_mN\right)}/\Delta_m .
\end{equation*}

In intermediate tumors, $N_a^* << N << N_m^*$, the sensitive cell population generates a number of rescue lineages in a short period of time after $\tau_a$. The time it takes the rescue lineages to reach the initial population size $N$ is small compared to the time it took the lineages to appear. Therefore, the mean recurrence time can be approximated by substituting \cref{meantimet2} for $\tau_a$ in \cref{meanproliferationtime} (black line in \Cref{ProliferationTimeLarge}).

In large tumors, $N_m^* \ll N$, the sensitive population produces a large number of rescue lineages in a short period of time.
As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}
